Cross-Loop Optimization of Arithmetic Intensity for Finite Element Local Assembly

### 2014-09-09 TACO-paper-v2/coffee.tex
UTF8 input encoding
Section 5: copy editing

\usepackage{graphicx}
\usepackage{caption}
\usepackage{subfig}
\usepackage{tabulary}
\usepackage{lineno}

@@

We present two studies. Firstly, in Section~\ref{sec:analysis-code-transf}, we
analyze the impact of the proposed code transformations, and their interplay,
in three real-world representative equations. We quantify the speed ups achievable at
the level of the local assembly kernel, i.e. by excluding from the measurements
the other sections of the computation. Then, in Section~\ref{sec:full-app}, we
provide a full-application investigation. Here, we use another equation (so as to
bring the total number of examined problems to four) to demonstrate that COFFEE's
optimizations actually allow a significant reduction in the overall computation run-time
to be achieved, which practically motivates our research.

@@

\subsubsection{Choice and Properties of the Benchmarks}
\label{sec:real-examples}
Our code transformations were evaluated in three real-world problems
based on the following PDEs
\begin{enumerate}
\item Helmholtz
\item Advection-Diffusion
\item Burgers
\end{enumerate}
The three chosen benchmarks are \emph{real-life kernels} and comprise
the core differential operators in some of the most frequently
encountered finite element problems in scientific computing. This is
of crucial importance because distinct problems, possibly arising  in
completely different fields, may employ (subsets of) the same differential
operators of our benchmarks, which implies similarities
and redundant patterns in the generated code. Consequently, the proposed
code transformations have a domain of applicability that goes far beyond
that of the three analyzed equations.

The Helmholtz and Diffusion kernels are archetypal second order elliptic
operators. They are complete and unsimplified examples of the
operators occurring for diffusion and viscosity in fluids, and for
imposing pressure in compressible fluids. As such, they are both
extensively used in climate and ocean modeling. Very similar
operators, for which the same optimisations are expected to be equally
effective, apply to elasticity problems, which are at the base of
computational structural mechanics. The Burgers kernel is a typical
example of a first order hyperbolic conservation law, which occurs in
real applications whenever a quantity is transported by a fluid (the
momentum itself, in our case). We chose this particular kernel since
it applies to a vector-valued quantity, while the elliptic operators
apply to scalar quantities; this impacts the generated code, as
explained next. In this way, the operators we have selected are
characteristic of both the second and first order operators that
dominate in fluids and solids simulations, and a wide variety on local
assembly codes can be tested.

The benchmarks were written in UFL (code available at~\cite{ufl-code})
and executed over real unstructured meshes through Firedrake. The
Helmholtz code has already been shown in
Listing~\ref{code:helmholtz}. For Advection-Diffusion, the Diffusion
equation, which uses the same differential operators as Helmholtz, is
considered. In the Diffusion kernel code, the main differences with
respect to Helmholtz are the absence of the $Y$ array and the presence
of a few more constants for computing the element matrix. Burgers is a
non-linear problem employing differential operators different from
those of Helmholtz and relying on vector-valued quantities, which has
a major impact on the generated assembly code (see
Listing~\ref{code:burgers}), where a larger number of basis function
arrays ($X1$, $X2$, ...) and constants ($F0$, $F1$, ..., $K0$, $K1$,
...) are generated.

These problems were studied varying both the shape of mesh elements
and the polynomial order $p$ of the method, whereas the element
family, Lagrange, is fixed. As might be expected, the larger the
element shape and $p$, the larger the iteration space. Triangles,
tetrahedron, and prisms were tested as element shape. For instance, in
the case of Helmholtz with $p=1$, the size of the \texttt{j} and
\texttt{k} loops for the three element shapes is, respectively, $3$,
$4$, and $6$. Moving to bigger shapes has the effect of increasing the
number of basis function arrays, since, intuitively, the behaviour of
the equation has now to be approximated also along a third axis. On
the other hand, the polynomial order affects only the problem size
(the three loops \texttt{i}, \texttt{j}, and \texttt{k}, and, as a
consequence, the size of $X$ and $Y$ arrays). A range of polynomial
orders from $p=1$ to $p=4$ were tested; higher polynomial orders are
excluded from the study because of current Firedrake limitations. In
all these cases, the size of the element matrix rarely exceeds
30$\times$30, with a peak of 105$\times$105 in Burgers with prisms and
$p=4$.

\subsubsection{Loop interchange}
In the following, only results for the loop order \texttt{ijk} are
shown. For the considerations exposed in
Section~\ref{sec:pyop2-compiler}, loop interchanges having an inner
loop along \texttt{i} caused slow downs; also, interchanging
\texttt{j} and \texttt{k} loops while keeping \texttt{i} as outermost
loop did not provide any benefits.

@@

\begin{table*}[t]
\tbl{Performance improvement due to generalized loop-invariant code
  motion, for different element shapes (triangle, tetrahedron, prism)
  and polynomial orders ($p \in [1, 4]$), over the original
  non-optimized code, for the Helmholtz, Diffusion and Burgers
  problems.}{ \scriptsize \licmresultsnorms }
\label{table:perf-results-licm}
\end{table*}
Table~\ref{table:perf-results-licm} illustrates the performance
improvement obtained when \emph{licm} is applied. In general, the
speed ups are notable. The main reasons were anticipated in
Section~\ref{sec:code-transf-licm}: in the original code, 1)
sub-expressions invariant to outer loops are not automatically
hoisted, while 2) sub-expressions invariant to the innermost loop are
hoisted, but their execution is not auto-vectorized. These
observations come from inspection of assembly code generated by the
compiler.

The gain tends to grow with the computational cost of the kernels:
bigger loop nests (i.e., larger element shapes and polynomial orders)
usually benefit from the reduction in redundant computation, even
though extra memory for the temporary arrays is required. Some
discrepancies to this trend are due to a less effective
auto-vectorization. For instance, on the Sandy Bridge, the improvement
at $p=3$ is larger than that at $p=4$ because, in the latter case, the
size of the innermost loop is not a multiple of the vector length, and
a reminder scalar loop is introduced at compile time. Since the loop
nest is small, the cost of executing the extra scalar iterations can
have a significant impact. The reminder loop overhead is more
pronounced on the Xeon Phi, where the vector length is twice as big,
which leads to proportionally larger scalar reminder loops.

@@

Table~\ref{table:perf-results-licmap} shows the cumulative impact of
\emph{licm}, data alignment, and padding over the original code. In
the following, this version of the code is referred to as
\emph{licm-ap}. Padding, which avoids the introduction of a reminder
loop as described in Section~\ref{licm-impact}, as well as data
alignment, enhance the quality of auto-vectorization. Occasionally the
run-time of \emph{licm-ap} is close to that of \emph{licm}, since the
non-padded element matrix size is already a multiple of the vector
length. Rarely \emph{licm-ap} is slower than \emph{licm} (e.g. in
Burgers $p=3$ on the Sandy Bridge). One possible explanation is that
the number of aligned temporaries introduced by \emph{licm} is so
large to induce cache associativity conflicts.

@@

Application of \emph{op-vect} to the Burgers problem induces
significant slow downs due to the large number of temporary arrays
that need to be tiled, which exceeds the available logical registers
on the underlying architecture. Expression splitting can be used in
combination with \emph{op-vect} to alleviate this issue; this is
discussed in the next section.

@@

\subsubsection{Impact of Expression Splitting}
\label{sec:perf-results-split}
Expression splitting relieves the register pressure when the element
matrix evaluation needs to read from a large number of basis function
arrays. As detailed in Section~\ref{sec:expr-split}, in the price to
pay for this optimazion, there are an increased number of accesses to
the element matrix and, potentially, redundant computation. Similarly
to the analysis of vector-register tiling, we compare two versions:
the baseline, \emph{licm-ap}; and expression splitting on top of
\emph{licm-ap}, which, for simplicity, in the following is referred to
as \emph{split}.

For the Helmholtz and Diffusion kernels, in which only between 4 and 8
temporaries are read at every loop iteration, \texttt{split} tends to
slow down the computation, because of the aforementioned
drawbacks. Slow downs up to 1.4$\times$ and up to 1.6$\times$ were
observed, respectively, on the Sandy Bridge and the Xeon Phi. Note
that the cost model prevents the adoption of the transformation: the
\texttt{while} statement in Figure~\ref{algo:applyCostModel} is indeed
never entered.

@@

@@

In these experiments, we observe a maximum performance improvement of 1.47$\times$ over the non-optimized local assembly code, obtained in the case ($f=2, p=2$). However, we reiterate the fact that full-application speed ups rise proportionally with the amount of time spent in assembly and, therefore, with the complexity of the equation. By increasing polynomial order and number of coefficient functions, or by simply studying a different, more complex equation, it is our experience that performance gains become increasingly more relevant. The choice of studying the static linear elasticity equation was to show that even relatively simple problems can be characterized by a large proportion of execution time spent in assembly.

>>>>

\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subfig}
\usepackage{tabulary}
\usepackage{lineno}

@@

We present two studies. Firstly, in
Section~\ref{sec:analysis-code-transf}, we analyze the impact of the
proposed code transformations, and their interplay, in three real-world
representative equations. We quantify the speedups achievable at the
level of the local assembly kernel, by excluding the other stages of the
computation from the measurements. Then, in Section~\ref{sec:full-app},
we provide a full-application investigation. Here, we use another
equation, bringing the total number of examined problems to four,
to demonstrate that COFFEE's optimizations actually allow a significant
reduction in the overall computation run-time to be achieved, which
practically motivates our research.

@@

\subsubsection{Choice and Properties of the Benchmarks}
\label{sec:real-examples}
Our code transformations were evaluated in three real-world problems
based on the following PDEs:
\begin{enumerate}
\item Helmholtz
\item Advection-Diffusion
\item Burgers
\end{enumerate}
The three chosen benchmarks are \emph{real-life kernels} and comprise
the core differential operators in some of the most frequently
encountered finite element problems in scientific computing. This is
of crucial importance because distinct problems, possibly arising  in
completely different fields, may employ (subsets of) the same differential
operators of our benchmarks, which implies similarities
and redundant patterns in the generated code. Consequently, the proposed
code transformations have a domain of applicability that goes far beyond
that of the three analyzed equations.

The Helmholtz and Diffusion kernels are archetypal second order elliptic
operators. They are complete and unsimplified examples of the
operators used to model diffusion and viscosity in fluids, and for
imposing pressure in compressible fluids. As such, they are both
extensively used in climate and ocean modeling. Very similar
operators, for which the same optimisations are expected to be equally
effective, apply to elasticity problems, which are at the base of
computational structural mechanics. The Burgers kernel is a typical
example of a first order hyperbolic conservation law, which occurs in
real applications whenever a quantity is transported by a fluid (the
momentum itself, in our case). We chose this particular kernel since
it applies to a vector-valued quantity, while the elliptic operators
apply to scalar quantities; this impacts the generated code, as
explained next. The operators we have selected are characteristic of
both the second and first order operators that dominate fluids and
solids simulations.
%and a wide variety on local assembly codes can be tested.

The benchmarks were written in UFL (code available at~\cite{ufl-code})
and executed over real unstructured meshes through Firedrake. The
Helmholtz code has already been shown in
Listing~\ref{code:helmholtz}. For Advection-Diffusion, the Diffusion
equation, which uses the same differential operators as Helmholtz, is
considered. In the Diffusion kernel code, the main differences with
respect to Helmholtz are the absence of the $Y$ array and the presence
of additional constants for computing the element matrix. Burgers is a
non-linear problem employing differential operators different from
those of Helmholtz and relying on vector-valued quantities, which has
a major impact on the generated assembly code (see
Listing~\ref{code:burgers}), where a larger number of basis function
arrays ($X1$, $X2$, ...) and constants ($F0$, $F1$, ..., $K0$, $K1$,
...) are generated.

These problems were studied varying both the shape of mesh elements
and the polynomial order $p$ of the method, whereas the element
family, Lagrange, is fixed. As might be expected, the larger the
element shape and $p$, the larger the iteration space. Triangles,
tetrahedra, and prisms were tested as element shape. For instance, in
the case of Helmholtz with $p=1$, the size of the \texttt{j} and
\texttt{k} loops for the three element shapes is, respectively, $3$,
$4$, and $6$. Moving to bigger shapes has the effect of increasing the
number of basis function arrays, since, intuitively, the behaviour of
the equation has to be approximated also along a third axis. On
the other hand, the polynomial order affects only the problem size
(the three loops \texttt{i}, \texttt{j}, and \texttt{k}, and, as a
consequence, the size of $X$ and $Y$ arrays). A range of polynomial
orders from $p=1$ to $p=4$ were tested; higher polynomial orders are
excluded from the study because of current Firedrake limitations. In
all these cases, the size of the element matrix rarely exceeds
30$\times$30, with a peak of 105$\times$105 in Burgers with prisms and
$p=4$.

\subsubsection{Loop interchange}
In the following, only results for the loop order \texttt{ijk} are
shown. For the considerations exposed in
Section~\ref{sec:pyop2-compiler}, loop interchanges having an inner
loop along \texttt{i} caused slowdowns; also, interchanging
\texttt{j} and \texttt{k} loops while keeping \texttt{i} as outermost
loop did not provide any benefits.

@@

\begin{table*}[t]
\tbl{Performance improvement due to generalized loop-invariant code
  motion, for different element shapes (triangle, tetrahedron, prism)
  and polynomial orders ($p \in [1, 4]$), over the original
  non-optimized code, for the Helmholtz, Diffusion and Burgers
  problems.}{ \scriptsize \licmresultsnorms }
\label{table:perf-results-licm}
\end{table*}
Table~\ref{table:perf-results-licm} illustrates the performance
improvement obtained when \emph{licm} is applied. In general, the
speedups are notable. The main reasons were anticipated in
Section~\ref{sec:code-transf-licm}: in the original code, 1)
sub-expressions invariant to outer loops are not automatically
hoisted, while 2) sub-expressions invariant to the innermost loop are
hoisted, but their execution is not auto-vectorized. These
observations come from inspection of assembly code generated by the
compiler.

The gain tends to grow with the computational cost of the kernels:
bigger loop nests (i.e., larger element shapes and polynomial orders)
usually benefit from the reduction in redundant computation, even
though extra memory for the temporary arrays is required. Some
discrepancies to this trend are due to a less effective
auto-vectorization. For instance, on the Sandy Bridge, the improvement
at $p=3$ is larger than that at $p=4$ because, in the latter case, the
size of the innermost loop is not a multiple of the vector length, and
a remainder scalar loop is introduced at compile time. Since the loop
nest is small, the cost of executing the extra scalar iterations can
have a significant impact. The remainder loop overhead is more
pronounced on the Xeon Phi, where the vector length is twice as long,
which leads to proportionally larger scalar remainder loops.

@@

Table~\ref{table:perf-results-licmap} shows the cumulative impact of
\emph{licm}, data alignment, and padding over the original code. In
the following, this version of the code is referred to as
\emph{licm-ap}. Padding, which avoids the introduction of a remainder
loop as described in Section~\ref{licm-impact}, as well as data
alignment, enhance the quality of auto-vectorization. Occasionally the
run-time of \emph{licm-ap} is close to that of \emph{licm}, since the
non-padded element matrix size is already a multiple of the vector
length. Rarely \emph{licm-ap} is slower than \emph{licm} (e.g. in
Burgers $p=3$ on the Sandy Bridge). One possible explanation is that
the number of aligned temporaries introduced by \emph{licm} is so
large to induce cache associativity conflicts.

@@

Application of \emph{op-vect} to the Burgers problem induces
significant slowdowns due to the large number of temporary arrays
that need to be tiled, which exceeds the available logical registers
on the underlying architecture. Expression splitting can be used in
combination with \emph{op-vect} to alleviate this issue; this is
discussed in the next section.

@@

\subsubsection{Impact of Expression Splitting}
\label{sec:perf-results-split} Expression splitting relieves the
register pressure when the element matrix evaluation needs to read from
a large number of basis function arrays. As detailed in
Section~\ref{sec:expr-split}, the price to pay for this optimazion is an
increased number of accesses to the element matrix and, potentially,
redundant computation. Similarly to the analysis of vector-register
tiling, we compare two versions: the baseline, \emph{licm-ap}, and
expression splitting on top of \emph{licm-ap}, which, for simplicity, in
the following is referred to as \emph{split}.

For the Helmholtz and Diffusion kernels, in which only between 4 and 8
temporaries are read at every loop iteration, \texttt{split} tends to
slow down the computation, because of the aforementioned
drawbacks. Slow downs up to 1.4$\times$ and up to 1.6$\times$ were
observed, respectively, on the Sandy Bridge and the Xeon Phi. Note
that the cost model prevents the adoption of the transformation: the
\texttt{while} statement in Figure~\ref{algo:applyCostModel} is never
entered.

@@

% FIXME: fastest quadrature or does that include tensor representation?

@@

In these experiments, we observe a maximum performance improvement of
1.47$\times$ over the non-optimized local assembly code, obtained in the
case ($f=2, p=2$). However, we reiterate the fact that full-application
speed ups increase proportionally with the amount of time spent in
assembly and, therefore, with the complexity of the equation. By
increasing polynomial order and number of coefficient functions, or by
simply studying a different, more complex equation, it is our experience
that performance gains become increasingly more relevant. The choice of
studying the static linear elasticity equation was to show that even
relatively simple problems can be characterized by a large proportion of
execution time spent in assembly.
